Spatial Generalized Linear Models in \proglang{R} Using \pkg{spmodel}

### 2025-08-22 inst/manuscript/manuscript.tex
some updates

\author{
Michael Dumelle~\orcidlink{0000-0002-3393-5529}\\United States\\
Environmental Protection Agency \And Jay M. Ver
Hoef~\orcidlink{0000-0003-4302-6895}\\Alaska Fisheries\\
Science Center \And Matt
Higham~\orcidlink{0009-0006-4217-625X}\\St.~Lawrence University
}
\title{Spatial Generalized Linear Models in {R} Using {spmodel}}

\Plainauthor{Michael Dumelle, Jay M. Ver Hoef, Matt Higham}
\Plaintitle{Spatial Generalized Linear Models in R Using spmodel}
\Shorttitle{SPGLMs in {R} Using {spmodel}}

\Abstract{
Generalized linear models (GLMs) describe a non-normal response variable
that may be binary, count, skewed, or a proportion. Typically,
observations in a GLM are assumed independent of one another. For
spatial data, this independence assumption is impractical, as nearby
locations tend to be more similar than locations far apart. The
{spmodel} {R} package provides tools to fit GLMs that incorporate
spatial autocorrelation (i.e., spatial generalized linear models, or
SPGLMs). SPGLMs are fit in {spmodel} using a novel application of the
Laplace approximation via {spglm} for point-referenced data or
{spgautor} for areal (i.e., lattice), data. {spglm} and {spgautor}
closely resemble {glm} from base {R} but include arguments that control
the spatial autocorrelation structure. {spmodel} has many helper
functions for model inspection and diagnostics, some of which leverage
other {R} packages like {broom} and {emmeans}. {spmodel} has tools to
make predictions of the latent spatial-mean process at unobserved
locations. {spmodel} also provides many advanced features like
accommodating geometric anisotropy and nonspatial random effects,
simulating spatially autocorrelated data, and more. Here we use
{spmodel} to illustrate the modeling of binary, count, skewed and
proportion response variables from several point-referenced and areal
data sets.
}

\Keywords{autoregressive model, geostatistical model, spatial
autocovariance, spatial autocorrelation}
\Plainkeywords{autoregressive model, geostatistical model, spatial
autocovariance, spatial autocorrelation}

@@

Binary, count, proportion, and skewed data are ubiquitous in practice.
These data types are naturally modeled using a generalized linear model
(GLM) framework
\citep{nelder1972generalized, mccullagh1989generalized, myers2012generalized, faraway2016extending}.
In a GLM, a response variable \(\text{y}\) belongs to a particular
statistical distribution with mean \(\mu\). For example, \(\text{y}\)
may be distributed as a Poisson random variable with some mean \(\mu\).
Based on the response distribution assumed for \(\text{y}\), GLMs link a
function of \(\mu\) to explanatory variables via a link function:
\begin{equation}\label{eq-glm}
f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
\end{equation} where for a sample size \(n\), \(\boldsymbol{\mu}\) is
the mean of an \(n \times 1\) response vector \(\mathbf{y}\),
\(f(\boldsymbol{\mu})\) is the link function that connects
\(\boldsymbol{\mu}\) to \(\mathbf{w}\), \(\mathbf{X}\) is the
\(n \times p\) design matrix of explanatory variables, and
\(\boldsymbol{\beta}\) is the \(p \times 1\) vector of fixed effects.
The mean, \(\boldsymbol{\mu}\), is usually constrained in some way
(e.g., positive) that depends on the distribution assumed for
\(\mathbf{y}\), but \(\mathbf{w}\) is unconstrained. The parameters in
Equation\(~\)\eqref{eq-glm} are typically estimated via maximum
likelihood (e.g., iteratively reweighted least squares)
\citep{chambers1992S}. The {glm} function is commonly used to fit GLMs
in the {R} programming language \citep{rcore2024}.

The GLM framework in Equation\(~\)\eqref{eq-glm} assumes the elements of
\(\mathbf{y}\) are independent of one another. This assumption is
impractical for spatial data, where nearby observations tend to be more
similar than distant observations \citep{tobler1970computer}. Ignoring
this spatial dependence can give rise to misleading inference and poor
prediction \citep{zimmerman2024spatial}. Spatial GLMs (SPGLMs) formally
incorporate spatial autocorrelation into a GLM by adding to
Equation\(~\)\eqref{eq-glm} two random effects that elucidate spatial
structure: \begin{equation}\label{eq-spglm}
f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
\end{equation} where \(\boldsymbol{\tau}\) is an \(n \times 1\) column
vector of spatially dependent random errors, and
\(\boldsymbol{\epsilon}\) is an \(n \times 1\) column vector of
spatially independent random errors. We make a few assumptions about
\(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\): first, that
\(\text{E}(\boldsymbol{\tau}) = \text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}\),
where \(\text{E}(\cdot)\) denotes expectation; second, that
\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}\), where
\(\mathbf{R}\) is an \(n \times n\) matrix that determines the spatial
dependence structure in \(\mathbf{w}\) (which is linked to
\(\boldsymbol{\mu}\) and hence \(\mathbf{y}\)) and depends on a range
parameter, \(\phi\); third, that
\(\text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_\epsilon \mathbf{I}\),
where \(\mathbf{I}\) is an \(n \times n\) identity matrix; and fourth,
that \(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\) are independent
of one another. The parameter \(\sigma^2_{\tau}\) is called the
spatially dependent random error variance or partial sill. The parameter
\(\sigma^2_\epsilon\) is called the spatially independent random error
variance or nugget. These two variance parameters are henceforth more
intuitively written as \(\sigma^2_{de}\) and \(\sigma^2_{ie}\),
respectively. The covariance of \(\mathbf{w}\) is denoted
\(\boldsymbol{\Sigma}\) and is given by \begin{equation}\label{eq-spcov}
 \boldsymbol{\Sigma} = \sigma^2_{de}\mathbf{R} + \sigma^2_{ie} \mathbf{I}.
\end{equation} The parameters \(\sigma^2_{de}\), \(\phi\), and
\(\sigma^2_{ie}\) are elements of \(\boldsymbol{\theta}\), the
covariance parameter vector.

The {spmodel} {R} package provides tools for fitting spatial statistical
models and making predictions at unobserved locations
\citep{dumelle2023spmodel}. A recent major update (v0.4.0) to {spmodel}
added SPGLM (Equation\(~\)\ref{eq-spglm}) support for binary, count,
skewed, and proportion response variables (Table\(~\)\ref{table-links}),
greatly expanding the class of models that {spmodel} makes accessible to
practitioners.

\begin{table}
\centering
\begin{tabular}{||c|ccc||} 
 \hline
 Family & Link Function & Link Name & Data Type  \\ [0.5ex] 
 \hline \hline
 Binomial & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Binary; Binary Count \\
 Poisson & $f(\mu) = \log(\mu)$ & Log & Count \\
 Negative Binomial & $f(\mu) = \log(\mu)$ & Log & Count \\
 Beta & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Proportion \\
 Gamma & $f(\mu) = \log(\mu)$ & Log & Skewed \\
 Inverse Gaussian & $f(\mu) = \log(\mu)$ & Log & Skewed \\ 
 \hline\hline
\end{tabular}
\caption{SPGLM response distributions and their link functions and data types.}
\label{table-links}
\end{table}

SPGLMs for point-referenced data are called geostatistical GLMs and are
fit using {spglm}. Data are point-referenced when the elements in
\(\mathbf{y}\) are observed at point-locations indexed by x-coordinates
and y-coordinates on a spatially continuous surface with an infinite
number of locations (e.g., point locations in a field). SPGLMs for areal
data are called spatial autoregresive models and are fit using
{spgautor}. Data are areal (i.e., lattice) when they are part of a
finite network of polygons whose connections are indexed by a
neighborhood structure (e.g., states in a country).

Several other {R} packages exist for analyzing SPGLMs. The {brms}
\citep{burkner2017brms}, {carBayes} \citep{lee2013carbayes}, {ngspatial}
\citep{hughes2020ngspatial}, {R-INLA} \citep{lindgren2015bayesian},
{spBayes} \citep{finley2007spbayes}, and {spNNGP}
\citep{finley2002spnngp} packages take a Bayesian approach, either
directly sampling from posterior distributions of parameters (e.g.,
using MCMC) or approximating them. A benefit of Bayesian approaches is
that prior information can be incorporated and uncertainty
quantification of parameter estimates is straightforward. However,
Bayesian approaches, especially those using MCMC, tend to be
computationally expensive. In order to reduce computation time, many of
these packages work with the precision matrix instead of the covariance
matrix so that inverting matrices (a computational bottleneck) is not
required. For example, {R-INLA} uses the precision matrix (often
directly modeled as with autoregressive models) and tends to be very
fast. Working with precision matrices, however, can be more restrictive
and less intuitive than working directly with the covariance matrix. The
{FRK} \citep{sainsbury2024modeling}, {glmmTMB}
\citep{brooks2017glmmtmb}, {hglm} \citep{ronnegard2010hglm}, {mgcv}
\citep{wood2017generalized}, and {spaMM} \citep{rousset2014spamm}
packages directly use Laplace, quasi-likelihood, or reduced-rank
approaches to estimate parameters. These direct approaches tend to be
computationally efficient, as they don't rely on MCMC sampling. In
contrast to the Bayesian approach, a drawback of these direct approaches
is that prior information cannot be formally incorporated and covariance
parameter uncertainty is challenging to quantify. SPGLMs in {spmodel}
are most similar to those in {glmmTMB} -- {spmodel} uses analytical
solutions for maximizing the likelihood (which we describe in the next
section), while {glmmTMB} uses automatic differentiation for maximizing
the likelihood (which tends to be much slower than analytical
solutions).

Missing from the aforementioned {R} packages is the complete set of
tools for SPGLMs that {spmodel} provides. Importantly, the {spglm} and
{spautor} functions act as a spatial analogue to the familiar {glm} from
base {R}, making the transition from GLMs to SPGLMs relatively seamless.
{spmodel} leverages many commonly used {R} generics like {summary} to
better understand fitted models. Six GLM families
(Table\(~\)\ref{table-links}) and 20 spatial covariance (or precision)
functions are supported. Also available are functions for data
visualization, model fitting, model summaries, model diagnostics, model
comparison, and prediction, all crucial components of a data analysis.
Through extra function arguments, {spmodel} supports advanced features
like geometric anisotropy, non-spatial random effects, methods for large
data sets, and more. Importantly, {spmodel} extends popular packages
like {broom} \citep{robinson2021broom, kuhn2022tidy} and {emmeans}
\citep{lenth2024emmeans}. {spmodel} also provides function for
simulating SPGLM data (e.g., {sprbinom}, {sprpois}).

The rest of this article is organized as follows. In
Section\(~\)\ref{sec-spglm}, we provide some background for the SPGLM
fitting and prediction routines in {spmodel}. In
Section\(~\)\ref{sec-applications}, we provide several applications of
{spmodel} to spatial binary, count, skewed and proportion data with both
point-referenced and areal supports. And in
Section\(~\)\ref{sec-discussion}, we end with a discussion synthesizing
\pkg{spmodel}'s contributions to the analysis of SPGLMs in {R}.

\section{Spatial generalized linear models using the Laplace
approximation}\label{sec-spglm}

GLMs with random effects are often written hierarchically
\citep{lee1996hierarchical, bolker2009generalized, wood2017generalized}.
{spmodel} leverages this hierarchical structure and uses a novel
application of the Laplace approximation \citep{ver2024marginal} to fit
models. This marginal approach is quite flexible, accommodating a wide
range of possible dependence structures and formally maximizing a
likelihood. Maximizing a likelihood yields convenient likelihood-based
statistics like AIC \citep{akaike1974new}, AICc
\citep{hoeting2006model}, BIC \citep{schwarz1978estimating}, deviance
\citep{mccullagh1989generalized}, and likelihood ratio tests for model
comparison, a benefit compared to quasi-likelihood
\citep{wedderburn1974quasi, breslow1993approximate} or pseudo-likelihood
approaches \citep{wolfinger1993generalized}, which only specify the
first two moments of a distribution. \citet{ver2024marginal} provide
thorough context for the marginal approach and its associated details,
but next we provide an short overview of the methodology.

Our goal is to marginalize over the latent mean \(\mathbf{w}\) and fixed
effects \(\boldsymbol{\beta}\) in Equation\(~\)\eqref{eq-spglm} to
obtain a response distribution for \(\mathbf{y}\) that depends only on
the explanatory variables, \(\mathbf{X}\), the response distribution's
dispersion parameter, \(\varphi\), and the covariance parameters,
\(\boldsymbol{\theta}\). We can represent this marginal distribution
hierarchically as \begin{equation}\label{eq-marginal}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w} .
\end{equation} In Equation\(~\)\eqref{eq-marginal},
\([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]\) is the density for
\(\mathbf{y}\) (e.g., binomial, Poisson) given the latent mean and
dispersion parameter and
\([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\) is the Gaussian
density for \(\mathbf{w}\) given the explanatory variables, fixed
effects, and covariance parameters. Integrating \(\boldsymbol{\beta}\)
out of Equation\(~\)\eqref{eq-spglm} yields
\begin{equation}\label{eq-marginal2}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\mathbf{w},
\end{equation} where \([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\)
is the restricted (i.e., residual) Gaussian density
\citep[\citet{harville1977maximum},
\citet{wolfinger1994computing}]{patterson1971recovery} for
\(\mathbf{w}\) given the explanatory variables and covariance
parameters. This restricted Gaussian density is given by
\begin{equation}\label{eq-reml-def}
[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] = \frac{\exp(-\frac{1}{2}(\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}}) \boldsymbol{\Sigma}^{-1} (\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}})^T)}{(2 \pi)^{(n - p)/2} |\boldsymbol{\Sigma}|^{1/2}|\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X}|^{1/2}},
\end{equation} where
\(\tilde{\boldsymbol{\beta}} = (\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{w}\)
and \(|\cdot|\) denotes the determinant.

>>>>

\author{
Michael Dumelle~\orcidlink{0000-0002-3393-5529}\\United States\\
Environmental Protection Agency \And Jay M. Ver
Hoef~\orcidlink{0000-0003-4302-6895}\\Alaska Fisheries\\
Science Center \And Matt
Higham~\orcidlink{0009-0006-4217-625X}\\St.~Lawrence University
}
\title{Spatial Generalized Linear Models in \proglang{R} Using
\pkg{spmodel}}

\Plainauthor{Michael Dumelle, Jay M. Ver Hoef, Matt Higham}
\Plaintitle{Spatial Generalized Linear Models in R Using spmodel}
\Shorttitle{Spatial Generalized Linear Models in \proglang{R} Using
\pkg{spmodel}}

\Abstract{
Generalized linear models (GLMs) describe a non-normal response variable
that may be binary, count, skewed, or a proportion. Typically,
observations in a GLM are assumed independent of one another. For
spatial data, this independence assumption is impractical, as nearby
locations tend to be more similar than locations far apart. The
\pkg{spmodel} \proglang{R} package provides tools to fit GLMs that
incorporate spatial correlation (i.e., spatial generalized linear
models, or SPGLMs). SPGLMs are fit in \pkg{spmodel} using a novel
application of the Laplace approximation via \code{spglm()} for
point-referenced data or \code{spgautor()} for areal (i.e., lattice),
data. \code{spglm()} and \code{spgautor()} closely resemble {glm} from
base \proglang{R} but include arguments that control the spatial
correlation structure. \pkg{spmodel} has many helper functions for model
inspection and diagnostics, some of which leverage other \proglang{R}
packages like {broom} and {emmeans}. \pkg{spmodel} has tools to make
predictions of the latent spatial-mean process at unobserved locations.
\pkg{spmodel} also provides many advanced features like accommodating
geometric anisotropy and nonspatial random effects, simulating spatially
autocorrelated data, and more. Here we use \pkg{spmodel} to illustrate
the modeling of binary, count, skewed and proportion response variables
from several point-referenced and areal data sets.
}

\Keywords{autoregressive model, geostatistical model, spatial
covariance, spatial correlation}
\Plainkeywords{autoregressive model, geostatistical model, spatial
covariance, spatial correlation}

@@

In practice, non-Gaussian data are ubiquitous. Non-Gaussian that belong
to an exponential family data can be naturally modeled using a
generalized linear model (GLM) regression framework
\citep{nelder1972generalized, mccullagh1989generalized, myers2012generalized, faraway2016extending}.
In a GLM, an \(n \times 1\) response variable \(\mathbf{y}\) belongs to
a a statistical distribution (e.g., Poisson, Binomial) with some mean
and variance. Often, the analysis goal is to study the impact of a
linear function of several explanatory variables on \(\text{y}\) through
a GLM. In this context, the latent (i.e., unobserved) mean of
\(\mathbf{y}\), \(\boldsymbol{\mu}\), is linked to these explanatory
variables via a link function: \begin{equation}\label{eq-glm}
f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
\end{equation} where for a sample size \(n\), \(f(\cdot)\) is a link
function that connects \(\boldsymbol{\mu}\) to \(\mathbf{w}\),
\(\mathbf{X}\) is the \(n \times p\) design matrix of explanatory
variables, and \(\boldsymbol{\beta}\) is the \(p \times 1\) vector of
fixed effects. While the mean is typically constrained in some way
(e.g., between zero and one if a probability), the link function
generally makes \(\mathbf{w}\) is unconstrained. Common link functions
inlude the log odds (i.e., logit) link for binary and proprtion data and
the log link count and skewed data. Equation\(~\)\ref{eq-glm} can also
be written in terms of the inverse link function, \(f^{-1}(\cdot)\):
\begin{equation}\label{eq-glm2}
\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta} \equiv f^{-1}(\mathbf{w}) = f^{-1}(\mathbf{X} \boldsymbol{\beta}),
\end{equation}

The GLM fixed effects (\(\boldsymbol{\beta}\)) are typically estimated
via maximum likelihood \citep{chambers1992S}. It is often convenient to
compute the maximum likelihood estimates using the iteratively
reweighted least squares (IRWLS) algorithm \citep{wood2017generalized},
which is the approach used by the \code{glm()} function in the
\proglang{R} programming language \citep{rcore2024}. GLMs add an
additional layer of complexity compared to linear regression models, as
the left-hand size of Equation\(~\)\ref{eq-glm} is a function of the
mean of \(\mathbf{y}\) rather than \(\mathbf{y}\) itself (as in linear
regression models).

The standard GLM assumes the elements of \(\mathbf{y}\) are independent.
This independence assumption is typically impractical for spatial data.
In spatial data, nearby observations tend to be more similar than
distant observations \citep{tobler1970computer}, leading to positive
spatial covariance among observations. The consequences of ignoring
spatial covariance in statistical models for spatial data can be severe
and include imprecise parameter estimates as well as misleading standard
errors that inflate Type-I error rates and decrease power
\citep{zimmerman2024spatial}.

An approach for handling spatial data using a GLM is to assume
\(\mathbf{w}\) has spatial covariance. This is achieved by adding to
Equation\(~\)\ref{eq-glm} two random effects, \(\boldsymbol{\tau}\) and
\(\boldsymbol{\epsilon}\). The random effect \(\boldsymbol{\tau}\) is an
\(n \times 1\) column vector of spatially dependent random errors. We
assume that \(\text{E}(\boldsymbol{\tau}) = \boldsymbol{0}\) and
\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}\), where
\(\text{E}(\cdot)\) and \(\text{Cov}(\cdot)\) denote expectation and
covariance, respectively. The variance parameter \(\sigma^2_\tau\)
controls the magnitude of spatial covariance and is often called a
partial sill, while the matrix \(\mathbf{R}\) is an \(n \times n\)
spatial correlation matrix that depends on a range parameter controls
the distance-decay rate of the spatial correlation. One example of a
spatial covariance matrix is the ``exponential'', which is given by
\begin{equation}\label{eq-spcov-exp}
  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{de} \exp(-\mathbf{H}/\phi),
\end{equation} where \(\mathbf{H}\) is a matrix of pairwise distances
among the elements of \(\mathbf{y}\) and \(\phi\) is a range parameter.
From Equation\(~\)\ref{eq-spcov-exp}, as the distance between two
elements of \(\mathbf{y}\) increases, the spatial covariance decreases,
which reflects intuition. Moreover, as the range parameter, \(\phi\),
increases, the strength of spatial dependence increases
(Figure\(~\)\ref{fig-range}). The random effect
\(\boldsymbol{\epsilon}\) is an \(n \times 1\) column vector of
independent random errors. We assume that
\(\text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}\) and
\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\epsilon \mathbf{I}\), where
\(\mathbf{I}\) is an \(n \times n\) identity matrix. The variance
parameter \(\sigma^2_\epsilon\) controls the magnitude of nonspatial
variability (i.e., fine-scale variation) and is often called a nugget.

\begin{figure}
\centering
\includegraphics[width = 0.7\linewidth]{figures/figure-01.png}
\caption{An exponential spatial correlation function with varying range parameters.}
\label{fig-range}
\end{figure}

Through inclusion of \(\boldsymbol{\tau}\) and
\(\boldsymbol{\epsilon}\), the spatial GLM (SPGLM) can be written as
\begin{equation}\label{eq-spglm}
f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\tau}, \boldsymbol{\epsilon}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon}.
\end{equation} Often in spatial statistics, quantities are explicitly
referenced with respect to \(\mathbf{s}\), a vector of coordinates
indexing the observation \citep{cressie1993statistics}. For example,
\(\mathbf{y}\) and \(\mathbf{X}\) may instead be written
\(\mathbf{y}(\mathbf{s})\) and \(\mathbf{X}\), respectively. We
acknowledge the utility of this nomenclature but drop the explicit
dependence on \(\mathbf{s}\) for simplicity of notation. Assuming
independence among \(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\),
it follows that \begin{equation}\label{eq-spcov}
 \text{Cov}(\boldsymbol{\tau} + \boldsymbol{\epsilon}) = \text{Cov}(\boldsymbol{\tau}) + \text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_{\tau}\mathbf{R} + \sigma^2_{\epsilon} \mathbf{I}.
\end{equation} To better align with intuition, we henceforth
\(\sigma^2_{\tau}\) as \(\sigma^2_{de}\) (for spatial error variance)
and \(\sigma^2_{\epsilon}\) as \(\sigma^2_{ie}\) (for independent error
variance). The parameters \(\sigma^2_{de}\), \(\sigma^2_{ie}\), the
range parameter \(\phi\) in \(\mathbf{R}\), and any other parameters in
\(\mathbf{R}\) compose \(\boldsymbol{\theta}\), the covariance parameter
vector.

Fitting and using SPGLMs is challenging both conceptually and
computationally \citep{bolker2009generalized}. Recently, however, there
have been numerous, significant advances in \proglang{R} software that
have made these models more accessible to practitioners. The \pkg{brms}
\citep{burkner2017brms}, \pkg{carBayes} \citep{lee2013carbayes},
\pkg{ngspatial} \citep{hughes2020ngspatial}, \pkg{R-INLA}
\citep{lindgren2015bayesian} and \pkg{inlabru} \citep{bachl2019inlabru},
\pkg{spBayes} \citep{finley2007spbayes}, \pkg{spOccupancy}
\citep{doser2022spoccupancy}, \pkg{spAbundance}
\citep{doser2024spabundance}, and \pkg{spNNGP} \citep{finley2002spnngp}
packages take a Bayesian approach, either directly sampling from
posterior distributions of parameters (e.g., using MCMC) or
approximating them. A benefit of Bayesian approaches is that prior
information can be incorporated and uncertainty quantification of
parameter estimates is straightforward. However, Bayesian approaches,
especially those using MCMC, can be computationally expensive. In order
to reduce computation time, many of these packages work with the
precision matrix instead of the covariance matrix so that
computationally expensive matrix inversion is not required. For example,
\pkg{R-INLA} uses the precision matrix and tends to be very fast.
Working with precision matrices, however, can be more restrictive and
less intuitive than working directly with the covariance matrix. The
{FRK} \citep{sainsbury2024modeling}, \pkg{glmmTMB}
\citep{brooks2017glmmtmb}, \pkg{hglm} \citep{ronnegard2010hglm},
\pkg{mgcv} \citep{wood2017generalized}, and \pkg{spaMM}
\citep{rousset2014spamm} packages directly use Laplace,
quasi-likelihood, or reduced-rank approaches to estimate parameters.
These direct approaches tend to be computationally efficient, as they
don't rely on MCMC sampling. In contrast to the Bayesian approach, a
drawback of these direct approaches is that prior information cannot be
formally incorporated and covariance parameter uncertainty is more
challenging to quantify. The \pkg{sdmTMB} \citep{anderson2024sdmtmb}
package combines elements of \pkg{R-INLA}, \pkg{glmmTMB}, and properties
of Gaussian Markov random fields to fit a wide variety of SPGLMs, and
\pkg{tinyVAST} \citep{thorson2025tinyVAST} extends some of these models
to multivariate or (dynamic) structural equation models.

\citet{ver2024marginal} proposed a novel approach to fitting SPGLMs that
leverages the Laplace approximation while marginalizing over both the
latent \(\mathbf{w}\) and the fixed effects (\(\boldsymbol{\beta}\)) and
accommodating spatial covariance. \citet{ver2024marginal} showed that
this approach performed efficiently in a variety of simulation settings,
generally having appropriate confidence interval coverage for the fixed
effects and prediction interval coverage for new \(\mathbf{w}\). The
approach performed similarly to the Bayesian SPGLM approach in
\pkg{spBayes} and the automatic differentiation SPGLM approach in
\pkg{glmmTMB} but was much faster. At small sample sizes, the approach
outperformed the approximate Bayesian SPGLM approach in \pkg{R-INLA} and
had similar computational times. For moderate sample sizes, it performed
similarly to \pkg{R-INLA}, though \pkg{R-INLA} was faster. This novel
approach is particularly attractive for two reasons. First, it is
general enough that can be applied to any covariance structure (not just
spatial). Second, after estimating the covariance parameters, analytical
solutions exist for the fixed effects (and their standard errors) as
well as predictions of the latent \(\mathbf{w}\) at new locations (and
their standard errors). The \pkg{spmodel} \proglang{R} package
\citep{dumelle2023spmodel} recently provided full support for the
methods in \citet{ver2024marginal} applied to binary, count, skewed, and
proportion data for over 20 different spatial covariance types.

The \pkg{spmodel} \proglang{R} package \citep{dumelle2023spmodel}
recently provided a full set of modeling tools for SPGLMs fit using the
methods described in \citet{ver2024marginal}. These modeling tools are
approachable and mirror the familiar \code{glm()} syntax from
base-\proglang{R}, making the transition from GLMs to SPGLMs relatively
seamless. The \code{spglm()} function fits SPGLMs for point-referenced
data (e.g., x-coordinates and y-coordinates representing point locations
in a field), while the \code{spgautor()} function fits SPGLMs for areal
data (e.g., polygon boundaries representing geographic subsets of a
region). \pkg{spmodel} supports the binomial distribution for binary
data, Poisson and negative binomial distributions for count data, Gamma
and inverse Gaussian distributions for skewed data, and the beta
distribution for proportion data. There are 20 different spatial
covariance structures available including the exponential, Gaussian, and
spherical for point-referenced data (Figure\(~\)\ref{fig-type}) and the
conditional autoregressive, and simultaneous autoregressive structures
for areal data. \pkg{spmodel} provides tools for commonly used model
summaries, visualizations, and diagnostics (e.g., Cook's distance) using
standard \proglang{R} helper functions like \code{summary()},
\code{plot()}, and \code{cooks.distance()}. \pkg{spmodel} also provides
tools to predict \(\mathbf{w}\) at new locations and quantify
uncertainty in those prediction using \code{predict()}. This core
functionality, combined with several advanced features we describe
throughout the manuscript, enable \pkg{spmodel} to provide some novel
and important capabilities previously missing from the existing SPGLM
ecosystem in \proglang{R}.

\begin{figure}
\centering
\includegraphics[width = 0.7\linewidth]{figures/figure-02.png}
\caption{Exponential, Gaussian, and spherical spatial correlation functions all with range parameters equal to 0.5.}
\label{fig-type}
\end{figure}

\pkg{spmodel} (version 0.11.0) is arguably most similar to \pkg{sdmTMB}
(version 0.7.4) in terms of scope and feel. Both packages use similar
syntax as \code{glm()}, accommodate flexible \code{formula} arguments
(e.g., offsets, splines), handle spatial covariance that decays at
different rates in different rates (i.e., geometric anisotropy),
incorporate nonspatial random effects, support other \proglang{R}
packages for modeling like \pkg{broom}
\citep{robinson2021broom, kuhn2022tidy} and \pkg{emmeans}
\citep{lenth2024emmeans}, and have tools for model summaries,
prediction, and simulating data. There are some notable differences
between the two packages, however. \pkg{sdmTMB} supports several
additional GLM distributions like the Tweedie, supports Hurdle models,
and can incorporate prior information through Bayesian applications.
\pkg{sdmTMB} also provides tools for working with temporal data and
enhanced visualizations of marginal effects. \pkg{sdmTMB} does require a
preprocessing step of constructing a mesh for the stochastic partial
differential equation approach, and the density of the mesh can affect
model results and computational complexity. \pkg{spmodel} does not
require the construction of a mesh prior to modeling. \pkg{spmodel}
supports 20 different spatial covariances and models them directly,
rather than using a precision matrix approximation to the Matérn spatial
covariance as in \pkg{sdmTMB}. \pkg{spmodel} also provides experimental
design tools (e.g., analysis of variance, contrasts), supports \pkg{sf}
objects in modeling and prediction functions \citep{pebesma2018sf}, has
several specialized model diagnostics like leverage values and Cook's
distances, and has analytic solutions for prediction standard errors.
Other similarities and differences do exist between \pkg{sdmTMB} and
\pkg{spmodel}, and both packages continue to evolve. Overall, we believe
that these packages are complementary and enhance the suite of SPGLM
tools accessible to practitioners.

The rest of this article is organized as follows. In
Section\(~\)\ref{sec-spglm}, we provide some background for the SPGLM
fitting and prediction routines in \pkg{spmodel}. In
Section\(~\)\ref{sec-applications}, we provide several applications of
\pkg{spmodel} to spatial binary, count, skewed and proportion data with
both point-referenced and areal supports. And in
Section\(~\)\ref{sec-discussion}, we end with a discussion synthesizing
\pkg{spmodel}'s contributions to the analysis of SPGLMs in \proglang{R}.

\section{The spatial generalized linear model and
marginalizatoin}\label{sec-spglm}

\pkg{spmodel} implements the novel methods described in
\citet{ver2024marginal} to fit SPGLMs, which leverages the Laplace
approximation and marginalizes over both the latent \(\mathbf{w}\) and
the fixed effects while accommodating spatial covariance. A beneficial
aspect of this approach is that it formally maximizes a hierarchical GLM
likelihood \citep{lee1996hierarchical, wood2017generalized}. This makes
likelihood-based statistics for model comparison like AIC
\citep{akaike1974new}, AICc \citep{hoeting2006model}, BIC
\citep{schwarz1978estimating}, deviance
\citep{mccullagh1989generalized}, and likelihood ratio tests available.
These types of statistics are not available for quasi-likelihood
\citep{wedderburn1974quasi, breslow1993approximate} or pseudo-likelihood
approaches \citep{wolfinger1993generalized}, which only specify the
first two moments of a distribution. \citet{ver2024marginal} provides
thorough details regarding the method and contextualizes its development
which built upon similar methods
\citep[\citet{bonat2016practical}]{evangelou2011estimation}. Next, we
describe a brief overview of the approach and how it can be used for
parameter estimation, inference, and prediction.

\subsection{Formulating the hierarchical
likelihood}\label{formulating-the-hierarchical-likelihood}

We can write the SPGLM likelihood hierarchically as
\begin{equation}\label{eq-marginal}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w},
\end{equation} where \([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]\) is
the density for the appropriate response distribution of \(\mathbf{y}\)
(e.g., binomial, Poisson) given the latent \(\mathbf{w}\) and dispersion
parameter (\(\varphi\)), and
\([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\) is the multivariate
Gaussian density for \(\mathbf{w}\) given the explanatory variables
(\(\mathbf{X}\)), fixed effects (\(\boldsymbol{\beta}\)), and spatial
covariance parameters (\(\boldsymbol{\theta}\)). Following
@\citet{harville1977maximum}, we can integrate \(\boldsymbol{\beta}\)
out of Equation\(~\)\ref{eq-spglm}, which yields
\begin{equation}\label{eq-marginal2}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\mathbf{w},
\end{equation} where \([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\)
is the restricted (i.e., residual) multivariate Gaussian density
\citep{patterson1971recovery} for \(\mathbf{w}\) given the explanatory
variables and covariance parameters. Equation\(~\)\ref{eq-marginal2} can
be synonymous written after profiling the overall variance out of
